Scalable and Modular Verified Compilation via Automated Proofs for Declarative Transformations

### 2025-01-31 paper.tex
remove dot from title
add progress report & conclusion

%% Title information
% \title{First Year Report}       %% [Short Title] is optional;
                                      %% when present, will be used in
                                      %% header instead of Full Title.
\title{Scalable and Modular Verified Compilation via Automated Proofs
for Declarative
Transformations.}                                      
\subtitle{First Year Report}

@@

\subsubsection*{Verified Compilation and Lean's Extensibility}

Note that the benefits of a verified compilation pipeline go far beyond
simply reducing the chance of bugs. It could lead to a culture
shift in the way software is written. Traditionally, writing compiler
optimizations was a domain reserved for the wise elders who we could
trust to perform the appropriate ``black magic''. Reasoning about the
correctness of a transformation was simply too hard, so we had to rely
on interpersonal trust. Tools like Alive have already caused a shift in
this approach: nowadays PRs that propose adding a new peephole
optimization to LLVM can include a link to Alive's online interface
showing the correctness, meaning no further discussion or review is
required. However, this effect is still limited to making reviews for
proposed additions to LLVM easier.

@@

>>>>

%% Title information
% \title{First Year Report}       %% [Short Title] is optional;
                                      %% when present, will be used in
                                      %% header instead of Full Title.
\title{Scalable and Modular Verified Compilation via Automated Proofs
for Declarative
Transformations}
\subtitle{First Year Report}

@@

@@

\section{Progress Report}\label{progress-report}

So far, I've worked on laying the foundations for my proposed research, in the form of the Lean-MLIR framework and bitvector decision procedures in Lean.

\subsection{Lean-MLIR}\label{lean-mlir}

As mentioned before, Lean-MLIR is a framework for giving semantics to MLIR dialects in Lean.
On top of this framework, we've built a peephole rewriter, and proved
that if a rewrite locally preserves semantics, then applying this
rewrite in the context of a larger program will preserve the semantics
of the program as a whole.

However, we've focused on dialects with pure (i.e., side effect free)
semantics. Even though we \emph{are} able to verify a lot of interesting
rewrites in high-level MLIR dialects, we do need to be able to reason
about side effect if we want to capture the low-level dialects as well.

To this end, I've\footnote{The framework is a collaborative effort; the handling of side effect is one of my contributions.} generalized the framework so that a dialects
semantics may be given in a user-defined monad, and proven that applying
a semantics-preserving pure rewrite (i.e., where the operations being
rewritten are all side effect free) inside a side-effecting program
preserves semantics of the program as a whole.

\subsection{Bitvectors}\label{bitvectors}

As mentioned, a big reason for us to use Lean is its community. As
such, I've been heavily involved in the redesign of Lean's API for
bitvectors, and the move of bitvectors from Mathlib to become part of Lean's core library
(i.e., the library that is by default included in every Lean program).
By ensuring our representation of bitvectors is aligned with the rest of
the community, we ensure we can easily reap the benefits of, e.g.,
tactics to automate bitvector proofs made by others.

One such effort is the verified bitblaster and SAT solver integration written by the Lean FRO.
Siddharth (another PhD student in our group) and I have contributed to this project by 
formally proving equivalences between bitvector operations and bitblast-able circuits.

As the semantics of both LLVM IR and assembly heavily involves bitvectors, much of 
the automation I propose will be built on top of this bitvector decision procedure.

\subsection{Amazon Internship}\label{amazon-internship}

At Amazon, I wrote a scalable symbolic simulator for straight-line, memory-free Arm machine code, in the context of LNSym, 
a toolkit for the verification of cryptographic hash functions and block ciphers written in Arm assembly.
Before my internship, a prototype symbolic simulator existed, but it was limited in features and did not scale well.
My reimplemented version tracks more information of the simulated execution, 
and is capable of simulating more than 400 instructions of the SHA512 litmus test we benchmarked against before hitting Lean's default time-out limits.
This is about 10x more instructions than the prototype was able to simulate given the same time-out limits.

The prototype relied heavily on Lean's builtin simplifier. 
I replaced this with bespoke metacode, that was able to exploit the specific shape of the problem we were solving.
Firstly, I wrote preprocessing that allowed us to treat the operational semantic model, where operations are given semantics as functions that write to state, 
in an axiomatic-semantics way, by having theorems that characterizes operations in terms of reads from the resulting state.
Secondly, I wrote a bespoke data-structure and used it selectively reflects parts of the proof goal.
We explicitly decoupled this meta-code needed to scale symbolic simulation from the definition of the semantic model,
so that the model remains extensible even by people who are not experts in formal verification, which was an explicit design goal.

This project, like some of my other research, was also an experiment in the usability of Lean as a platform to verify software.
In particular, we generated very large proofs, and exposed, amongst others, a fundamental scalability problem in Lean:
The time Lean needs to assemble a proof after a tactic executes scales roughly by the length of the proof times the number of binders in the proof.
In the larger benchmarks, LNSym would generate proofs involving tens of thousands of binders,
resulting in the majority of time needed to check the proof was actually spent in this proof-assembling step.
We reported this to the Lean core team, who are actively working on a solution.
Furthermore, the semantic model of Arm machine code used relies heavily on bitvectors, thus, we also used the proof obligations resulting from the improved symbolic simulator to stress-test the bitvector decision procedure. I've reported these problems, and submitted patches for some, contributing to the greater vision of improving the usability of Lean.

\section{Conclusion}

Implementation a verified compilation pipeline as described in this proposal could 
have an impact far beyond simply reducing the chance of bugs. It could lead to a culture
shift in the way software is written. Traditionally, writing compiler
optimizations was a domain reserved for the wise elders who we could
trust to perform the appropriate ``black magic''. Reasoning about the
correctness of a transformation was simply too hard, so we had to rely
on interpersonal trust. Tools like Alive have already caused a shift in
this approach: nowadays PRs that propose adding a new peephole
optimization to LLVM can include a link to Alive's online interface
showing the correctness, meaning no further discussion or review is
required. However, this effect is still limited to making reviews for
proposed additions to LLVM easier.
